---
title: "Reproducible Report on COVID19 Data"
author: "Christopher Chery"
date: '2025-06-19'
output:
  pdf_document: default
  html_document: default
---

## Project Objective

Utilizing Johns Hopkins University datasets, I will analyze COVID-19 pandemic data to address the following questions: which US state's population was most affected by the virus and how did the United States' mortality rate compare globally? Additionally, I will employ an ARIMA model to forecast US COVID-19 deaths for the first quarter of 2023.

## Data Overview

First, I will import the necessary libraries and import the COVID19 and population data from the five JHU csv files.

```{r library, message=FALSE, warning=FALSE}
library("tidyverse")
library("dplyr")
library("lubridate")
library("forecast")
library("tseries")
```


```{r get_covid19_data, echo=FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_US.csv",
                "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in, file_names)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
```

```{r read in datasets, echo=FALSE}

us_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
us_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
global_population <- read_csv(uid_lookup_url)

us_cases
global_cases
us_deaths
global_deaths
global_population
```

## Tidy and Transfrom Data

```{r tidy_global_data}
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region',
                         Lat,
                         Long),
               names_to = "date",
               values_to = "cases") %>% 
  select(-c(Lat, Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State',
                         'Country/Region',
                         Lat,
                         Long),
               names_to = "date",
               values_to = "deaths") %>% 
  select(-c(Lat, Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
  filter(cases > 0) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

global
```

```{r tidy_us_data, message = FALSE}
us_cases <- us_cases %>%
  pivot_longer(cols = -c(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths <- us_deaths %>%
  pivot_longer(cols = -c(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us <- us_cases %>%
  full_join(us_deaths) %>%
  filter(cases > 0) %>%
  rename(County = "Admin2")

us
```

```{r global_population}
global <- global %>%
  left_join(global_population, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population)


global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```

Now that the tidying and transformations are complete, these final data sets can be used for analysis.

```{r view_data}
us
summary(us)

global
summary(global)
```
## Exploratory Data Analysis

### Objective #1

For my first objective of determining which US state was most affected by COVID-19, I will summarize cases, deaths, and population by each state and again by the total United States. I will also create variables for cases per million, deaths per million, and mortality rate.

```{r state_summary, message = FALSE}
state_pop <- us %>%
  distinct(Province_State, County, .keep_all = TRUE) %>%
  group_by(Province_State) %>%
  summarize(Population = sum(Population))

us_by_state <- us %>%
  group_by(Country_Region, Province_State, date) %>%
  summarize(cases = sum(cases),
            deaths = sum(deaths)) %>%
  ungroup() %>%

  left_join(state_pop, by = "Province_State") %>%
  filter(Population > 0) %>%
  filter(!is.na(Population)) %>%

  mutate(deaths_per_mill = deaths * 1000000 / Population,
         cases_per_mill = cases * 1000000 / Population,
         mortality_rate = deaths/ cases) %>%
  select(Province_State, date, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)
 
us_states_ovr <- us_by_state %>%
  group_by(Province_State) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  select(Province_State, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)
  
us_by_state
us_states_ovr
```

Now I will plot my Death per Million variable to identify the top 10 states that were most affected by the COVID-19 deaths.

```{r state_mort_rates}
top_10_states <- us_states_ovr %>%
  arrange(desc(deaths_per_mill)) %>%
  head(10)

ggplot(top_10_states, aes(x = deaths_per_mill, y = reorder(Province_State, deaths_per_mill))) +
  geom_point(color = "darkred", size = 3) +
  labs(title = "COVID-19 Deaths per Million by US State",
       subtitle = "Top 10",
       x = "Deaths per Million",
       y = "State") +
  theme_minimal()
```

The plot shows that relative to population, Arizona was the state most affected by COVID-19 deaths.

### Objective #2

For my second objective of determining how the US's mortality rate compares to the rest of the world, I will now perform the same summarizations and create the same variables, but instead grouping on a national level. I will have 2 data-frames, one containing time-series data and another with a cumulative total.

```{r us_summary, message = FALSE}
us_pop <- us %>%
  distinct(Country_Region, Province_State, County, .keep_all = TRUE) %>%
  group_by(Country_Region) %>%
  summarize(Population = sum(Population))

us_totals <- us %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases),
            deaths = sum(deaths)) %>%
  ungroup() %>%

  left_join(us_pop, by = "Country_Region") %>%
  filter(Population > 0) %>%
  filter(!is.na(Population)) %>%

  mutate(deaths_per_mill = deaths * 1000000 / Population,
         cases_per_mill = cases * 1000000 / Population,
         mortality_rate = deaths/ cases) %>%
  select(Country_Region, date, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)

us_ovr <- us_totals %>%
  group_by(Country_Region) %>%
  filter(date == max(date)) %>%
  ungroup() %>%
  select(Country_Region, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)

us_totals
us_ovr
```

The same data-frames will now be built using the global data.

```{r global_summary, message = FALSE}
global_pop <- global %>%
  distinct(Country_Region, Province_State, .keep_all = TRUE) %>%
  group_by(Country_Region) %>%
  summarize(Population = sum(Population))

global_totals <- global %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases),
            deaths = sum(deaths)) %>%
  ungroup() %>%
  
  left_join(global_pop, by = "Country_Region") %>%
  filter(Population > 0) %>%
  filter(!is.na(Population)) %>%

  mutate(deaths_per_mill = deaths * 1000000 / Population,
         cases_per_mill = cases * 1000000 / Population,
         mortality_rate = deaths/ cases) %>%
  select(Country_Region, date, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)

global_ovr <- global_totals %>%
  group_by(Country_Region) %>%
  filter(date == max(date),
         cases > 1) %>%
  ungroup() %>%
  select(Country_Region, cases, cases_per_mill, deaths, deaths_per_mill, mortality_rate, Population)

global_totals
global_ovr
```

Now that my data-frames are complete, I will merge them together so that the data can be plotted. Since there is a large number of different countries in this data, I will be using a density plot to compare the global COVID-19 mortality rates.

```{r global_mort_rates, warning = FALSE}
merged_data <- bind_rows(global_ovr, us_ovr)

ggplot(merged_data, aes(x = mortality_rate)) +
  geom_density(fill = "lightblue", alpha = 0.5) +
  geom_vline(data = subset(merged_data, Country_Region == "US"),
             aes(xintercept = mortality_rate),
             color = "red", size = 1.2, linetype = "dashed") +
  annotate("text",
           x = subset(merged_data, Country_Region == "US")$mortality_rate,
           y = Inf,
           label = "USA",
           vjust = 8,
           hjust = -.5,
           color = "red") +
  labs(title = "Global COVID-19 Mortality Rate Distribution",
       x = "Mortality Rate",
       y = "Density") +
  scale_x_log10() +
  theme_minimal()
```

The density plot shows that the US has a COVID-19 mortality rate slightly above 1%, which appears to be in line with the global average rate.

### Objective 3

For my third and final objective, I will feed the 'US Totals' data-frame into an ARIMA model to predict COVID-19 deaths during the first quarter of 2023. The model will be trained using the data from 2020-2022, and the predicted deaths will be compared to the actual deaths for the first quarter of 2023.

```{r model_deaths}
model_data <- us_totals %>%
  filter(deaths > 0) %>%
  select(date, deaths)

train_data <- model_data %>% filter(date < as.Date("2023-01-01"))
test_data <- model_data %>% filter(date >= as.Date("2023-01-01"))

ts_train <- ts(train_data$deaths, start = c(2020, 1), frequency = 365)

ts_test <- ts(test_data$deaths, start = c(2023, 1), frequency = 365)

diff_train <- diff(diff(ts_train))

adf.test(diff_train)

arima_model <- auto.arima(diff_train)
summary(arima_model)

forecasted <- forecast(arima_model, h = length(ts_test))

forecasted_differences <- as.numeric(forecasted$mean)
first_cumsum <- cumsum(forecasted_differences) + as.numeric(tail(diff(ts_train), n = 1))
original_scale_predictions <- cumsum(first_cumsum) + as.numeric(tail(ts_train, n = 1))

predicted_dates <- seq(
  from = as.Date("2023-01-01"),
  by = "day",
  length.out = length(original_scale_predictions)
)

actual_deaths <- model_data %>%
  filter(date <= as.Date("2022-12-31"))

comparison <- bind_rows(
  train_data %>% filter(date >= as.Date("2023-01-01")), 
  data.frame(date = predicted_dates, deaths = test_data$deaths, predicted_deaths = original_scale_predictions)) %>%
  filter(year(date) == 2023)

ggplot(comparison, aes(x = date)) +
  geom_line(aes(y = deaths, color = "Actual"), size = 1, na.rm = TRUE) +
  geom_line(aes(y = predicted_deaths, color = "Predicted"), size = 1, na.rm = TRUE) +
  scale_color_manual(values = c("Actual" = "black", "Predicted" = "red")) +
  labs(
    title = "Actual vs Predicted COVID-19 Deaths",
    x = "Date",
    y = "Daily Deaths",
    color = "Legend") +
  theme_minimal()
```

## Conclusion

Utilizing data sets from Johns Hopkins University, I successfully achieved all research objectives. However, it's crucial to acknowledge potential biases within the analysis. Numerous factors influence COVID-19 cases, deaths, and associated mortality rates. The provided data sets don't account for critical variables like government policy, vaccination rates, or the lag time between diagnosis and death. Within the United States, these variables varied significantly across states and cities. Globally, some countries implemented stringent COVID-19 policies, while others adopted a more relaxed approach. Therefore, when interpreting the results of this analysis, it's essential to remember that the data's inability to account for these variables renders the findings more exploratory than definitively factual.

## R Session Info

```{r echo=FALSE}
utils::sessionInfo()
```
